---
output: github_document
bibliography: vignettes/references.json
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# _superspreading_: estimating individual-level variation in transmission

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/superspreading/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/superspreading/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/epiverse-trace/superspreading/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/superspreading?branch=main)
<!-- badges: end -->

`{superspreading}` provides a set of functions to estimating and calculating
metrics to understand individual-level variation in transmission from
secondary cases.

`{superspreading}` is developed at the [Centre for the Mathematical Modelling of Infectious Diseases](https://www.lshtm.ac.uk/research/centres/centre-mathematical-modelling-infectious-diseases) at the [London School of Hygiene and Tropical Medicine](https://www.lshtm.ac.uk/) as part of [Epiverse-TRACE](https://data.org/initiatives/epiverse/).

## Installation

The easiest way to install the development version of `{superspreading}` from
[GitHub](https://github.com/) is to use the `{pak}` package:
  
  ``` r
# check whether {pak} is installed
if(!require("pak")) install.packages("pak")
pak::pak("epiverse-trace/superspreading")
```

## Quick start

```{r load-superspreading}
library(superspreading)
```

### Calculate the heterogeneity of transmission

Case study using data from early Ebola outbreak in Guinea in 2014, stratified by index and non-index cases, as in [Kucharski et al (EID, 2016)](https://doi.org/10.3201%2Feid2201.151410)

```{r, estim-superspreading}
# we use {fitdistrplus} to fit the models
library(fitdistrplus)

# load data on transmission from index and secondary cases for Ebola in 2014
# Source: Faye et al (https://doi.org/10.1016/S1473-3099(14)71075-8) & Althaus et al: https://doi.org/10.1016/S1473-3099(15)70135-0

# total cases
n_all <- 152 
index_case_transmission <- c(2, 17, 5, 1, 8, 2, 14)
secondary_case_transmission <- c(
  1, 2, 1, 4, 4, 1, 3, 3, 1, 1, 4, 9, 9, 1, 2, 1, 1, 1, 4, 3, 3, 4, 2, 5, 
  1, 2, 2, 1, 9, 1, 3, 1, 2, 1, 1, 2
) 

# Format data into index and non-index cases
# total non-index cases
n_non_index <- sum(c(index_case_transmission,secondary_case_transmission)) 
# transmission from all non-index cases
non_index_cases <- c(
  secondary_case_transmission, 
  rep(0, n_non_index-length(secondary_case_transmission))
) 

# Estimate R and k for index and non-index cases
param_index <- fitdist(data = index_cases, distr = "nbinom") 
param_non_index <- fitdist(data = non_index_cases, distr = "nbinom") 
```

### Calculate the probability of large epidemic

```{r, prob-epidemic}
# Compare probability of a large outbreak when k varies according to index/non-index values, 
# assuming R=2 and 3 initial spillover infections

R_input = 2
initial_infections = 3

# Probability of epidemic using k estimate from index cases
probability_epidemic(R = R_input, k = param_index$estimate[["size"]],a = initial_infections)

# Probability of epidemic using k estimate from non-index cases
probability_epidemic(R = R_input, k = param_non_index$estimate[["size"]],a = initial_infections)
```

## Help 

To report a bug please open an [issue](https://github.com/epiverse-trace/superspreading/issues/new/choose)

## Contribute 

Contributions to `{superspreading}` are welcomed. Please follow the [package contributing guide](https://github.com/epiverse-trace/superspreading/blob/main/.github/CONTRIBUTING.md).

## Code of Conduct

Please note that the `{epiparameter}` project is released with a [Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

## Citing this package

```{r message=FALSE, warning=FALSE}
citation("superspreading")
```

### Code of Conduct
Please note that the {superspreading} project is released with a 
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
