---
output: github_document
bibliography: vignettes/references.json
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# _superspreading_: Estimate individual-level variation in transmission <img src="man/figures/logo.svg" align="right" width="120"/>

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/license/mit)
[![R-CMD-check](https://github.com/epiverse-trace/superspreading/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/superspreading/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/epiverse-trace/superspreading/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/superspreading?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/superspreading)](https://CRAN.R-project.org/package=superspreading)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/superspreading)](https://cran.r-project.org/package=superspreading)
<!-- badges: end -->

`{superspreading}` is an R package that provides a set of functions to estimate
and understand individual-level variation in transmission of infectious diseases
from data on secondary cases.

`{superspreading}` implements methods outlined in @antiaRoleEvolutionEmergence2003, @lloyd-smithSuperspreadingEffectIndividual2005, @kucharskiEarlyDynamicsTransmission2020, and @kremerQuantifyingSuperspreadingCOVID192021, as well as additional functions.

`{superspreading}` is developed at the [Centre for the Mathematical Modelling of Infectious Diseases](https://www.lshtm.ac.uk/research/centres/centre-mathematical-modelling-infectious-diseases) at the [London School of Hygiene and Tropical Medicine](https://www.lshtm.ac.uk/) as part of [Epiverse-TRACE](https://data.org/initiatives/epiverse/).

## Installation

The package can be installed from CRAN using

```r
install.packages("superspreading")
```

The easiest way to install the development version of `{superspreading}` from
[GitHub](https://github.com/) is to use the `{pak}` package:

``` r
# check whether {pak} is installed
if(!require("pak")) install.packages("pak")
pak::pak("epiverse-trace/superspreading")
```

Alternatively, install pre-compiled binaries from [the Epiverse TRACE R-universe](https://epiverse-trace.r-universe.dev/superspreading)

``` r
install.packages("superspreading", repos = c("https://epiverse-trace.r-universe.dev", "https://cloud.r-project.org"))
```

## Quick start

```{r load-superspreading}
library(superspreading)
```

### Calculate the heterogeneity of transmission

Case study using data from early Ebola outbreak in Guinea in 2014, stratified by index and non-index cases, as in @kucharskiEffectivenessRingVaccination2016. Data on transmission from index and secondary cases for Ebola in 2014.

***Source***: @fayeChainsTransmissionControl2015 & @althausEbolaSuperspreading2015.

[`{fitdistrplus}`](https://CRAN.R-project.org/package=fitdistrplus) is a well-developed and stable R package that provides a variety of methods for fitting distribution models to data [@delignette-mullerFitdistrplusPackageFitting2015]. Therefore, it is used throughout the documentation of `{superspreading}` and is a recommended package for those wanting to fit distribution models, for example those supplied in `{superspreading}` (Poisson-lognormal and Poisson-Weibull). We recommend reading the `{fitdistrplus}` documentation (specifically `?fitdist`) to explore the full range of functionality.

In this example we fit the negative binomial distribution to estimate the reproduction number ($R$, which is the mean of the distribution) and the dispersion ($k$, which a measure of the variance of the distribution). The parameters are estimated via maximum likelihood (the default method for `fitdist()`).

```{r, estim-superspreading}
# we use {fitdistrplus} to fit the models
library(fitdistrplus)

# transmission events from index cases
index_case_transmission <- c(2, 17, 5, 1, 8, 2, 14)

# transmission events from secondary cases
secondary_case_transmission <- c(
  1, 2, 1, 4, 4, 1, 3, 3, 1, 1, 4, 9, 9, 1, 2, 1, 1, 1, 4, 3, 3, 4, 2, 5,
  1, 2, 2, 1, 9, 1, 3, 1, 2, 1, 1, 2
)

# Format data into index and non-index cases
# total non-index cases
n_non_index <- sum(c(index_case_transmission, secondary_case_transmission))
# transmission from all non-index cases
non_index_cases <- c(
  secondary_case_transmission,
  rep(0, n_non_index - length(secondary_case_transmission))
)

# Estimate R and k for index and non-index cases
param_index <- fitdist(data = index_case_transmission, distr = "nbinom")
# rename size and mu to k and R
names(param_index$estimate) <- c("k", "R")
param_index$estimate
param_non_index <- fitdist(data = non_index_cases, distr = "nbinom")
# rename size and mu to k and R
names(param_non_index$estimate) <- c("k", "R")
param_non_index$estimate
```

The reproduction number ($R$) is higher for index cases than for non-index cases, but the heterogeneity in transmission is higher for non-index cases (i.e. $k$ is lower).

### Calculate the probability of a large epidemic

Given the reproduction number ($R$) and the dispersion ($k$), the probability that
a infectious disease will cause an epidemic, in other words the probability it
does not go extinct, can be calculated using `probability_epidemic()`. Here we
use `probability_epidemic()` for the parameters estimated in the above section
for Ebola, assuming there are three initial infections seeding the potential outbreak.

```{r, prob-epidemic}
# Compare probability of a large outbreak when k varies according to
# index/non-index values, assuming 3 initial spillover infections

initial_infections <- 3

# Probability of an epidemic using k estimated from index cases
probability_epidemic(
  R = param_index$estimate[["R"]],
  k = param_index$estimate[["k"]],
  num_init_infect = initial_infections
)

# Probability of an epidemic using k estimated from non-index cases
probability_epidemic(
  R = param_non_index$estimate[["R"]],
  k = param_non_index$estimate[["k"]],
  num_init_infect = initial_infections
)
```

The probability of causing a sustained outbreak is high for the index cases, but
is zero for non-index cases (i.e. disease transmission will inevitably cease assuming transmission
dynamics do not change).

## Package vignettes

More details on how to use `{superspreading}` can be found in the [online documentation as package vignettes](https://epiverse-trace.github.io/superspreading/), under "Articles".

### Visualisation and plotting functionality

`{superspreading}` does not provide plotting functions, instead we provide example code chunks in the package's vignettes that can be used as a templates upon which data visualisations can be modified. We recommend users copy and edit the examples for their own purposes. (This is permitted under the package's MIT license). By default code chunks for plotting are folded, in order to unfold them and see the code simply click the code button at the top left of the plot.

## Help

To report a bug please open an [issue](https://github.com/epiverse-trace/superspreading/issues/new/choose)

## Contribute

Contributions to `{superspreading}` are welcomed. Please follow the [package contributing guide](https://github.com/epiverse-trace/.github/blob/main/CONTRIBUTING.md).

## Code of Conduct

Please note that the {superspreading} project is released with a
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

## Citing this package

```{r message=FALSE, warning=FALSE}
citation("superspreading")
```

## Related projects

This project has some overlap with other R packages:

- [`{epichains}`](https://github.com/epiverse-trace/epichains) is another Epiverse-TRACE R package that analyses transmission chain data to infer the likelihood for either the size or length of an outbreak cluster, or simulate transmission chains. It is based on the, now retired, [`{bpmodels}`](https://github.com/epiforecasts/bpmodels) package.

    Two main differences between `{superspreading}` and `{epichains}` are: 1) `{superspreading}` has functions to compute metrics that characterise outbreaks and superspreading events (e.g. `probability_epidemic()`, `probability_extinct()`, `proportion_cluster_size()` & `proportion_transmission()`); whereas `{epichains}` has functions to calculate the likelihood of a transmission chain size and length. 2) `{epichains}` exports functions to simulate a single-type branching process (`simulate_chains()` & `simulate_chain_stats()`).

- [`{modelSSE}`](https://CRAN.R-project.org/package=modelSSE) has a similar scope to `{superspreading}`, it contains functions to infer offspring distribution parameters. It exports several infectious disease outbreak datasets (see `data(package = "modelSSE")`). Both `{superspreading}` and `{modelSSE}` export functions to calculate the proportion of transmission using different methods. It also imports the [`{Delaporte}`](https://CRAN.R-project.org/package=Delaporte) package to model the offspring distribution as a Delaporte distribution.

## References
