---
title: "Pathogen evolution in the emergence of infectious disease outbreaks"
output: rmarkdown::html_vignette
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pathogen evolution in the emergence of infectious disease outbreaks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

Infectious diseases that cannot transmit effectively between humans cannot cause sustained disease outbreaks. However, they can cause stuttering chains of transmission with several cases. These short human-to-human transmission chains provide opportunities for the pathogen to adapt to humans and enhance transmission potential. Therefore, even in cases where a pathogen initially introduced into a human population might have a reproduction number below 1, that is on average it is transmitted to less than one other person (bound to inevitably going extinct without pathogen evolution), if a pathogen mutates while infecting humans, its reproduction number might evolve to exceed 1, and emerge to cause sustained outbreaks. 

This vignette explores the `probability_emergence()` function. This is based on the framework of @antiaRoleEvolutionEmergence2003 which uses a multi-type branching process to calculate the probability that a subcritical zoonotic spillover will evolve into a more transmissible pathogen with a reproduction number greater than 1 and thus be able to cause a sustained epidemic from human-to-human transmission. 

```{r setup}
library(superspreading)
library(ggplot2)
library(scales)
```

We use `probability_emergence()` to replicate Figure 2b from @antiaRoleEvolutionEmergence2003 showing how the probability of emergence changes for different wild-type reproduction numbers and different mutation rates. This assumes a 2-type branching process with a wild-type pathogen with $R_0 < 1$ and a mutant pathogen with $R_0 > 1$, we test with mutant reproduction numbers of 1.2, 1.5, 1,000.

```{r calc-prob-emerge}
R_wild <- seq(0, 1.2, by = 0.01)
R_mutant <- c(1.2, 1.5, 1000)
mutation_rate <- c(10^-1, 10^-3)

params <- expand.grid(
  R_wild = R_wild,
  R_mutant = R_mutant,
  mutation_rate = mutation_rate
)

prob_emerge <- apply(
  params,
  MARGIN = 1,
  FUN = function(x) {
    probability_emergence(
      R_wild = x[["R_wild"]],
      R_mutant = x[["R_mutant"]],
      mutation_rate = x[["mutation_rate"]],
      num_init_infect = 1
    )
  }
)
res <- cbind(params, prob_emerge = prob_emerge)
```

```{r plot-prob-emerge}
#| fig.alt: >
#|   Line plot showing the probability of emergence of a pathogen as a function
#|   of the basic reproduction number (R0) of the introduced pathogen. The
#|   x-axis represents the R0 of the introduced pathogen, ranging from 0 to
#|   1.2. The y-axis shows the probability of emergence on a logarithmic scale
#|   from 1e-05 to 1. Lines vary by mutation rate (green for 0.001 and black
#|   for 0.1) and by mutant pathogen R0 (dotted for R0 = 1.2, dashed for R0 =
#|   1.5, solid for R0 = 1000). For both mutation rates, the probability of
#|   emergence increases with the R0 of the introduced pathogen and with the
#|   R0 of the mutant. Higher mutation rates and higher mutant R0s lead to
#|   higher probabilities of emergence.
ggplot(data = res) +
  geom_line(
    mapping = aes(
      x = R_wild,
      y = prob_emerge,
      colour = as.factor(mutation_rate),
      linetype = as.factor(R_mutant)
    )
  ) +
  scale_y_continuous(
    name = "Probability of emergence",
    transform = "log",
    breaks = log_breaks(n = 6),
    limits = c(10^-5, 1)
  ) +
  scale_colour_manual(
    name = "Mutation rate",
    values = c("#228B22", "black")
  ) +
  scale_linetype_manual(
    name = "Mutant pathogen R0",
    values = c("dotted", "dashed", "solid")
  ) +
  scale_x_continuous(
    name = expression(R[0] ~ of ~ introduced ~ pathogen),
    limits = c(0, 1.2),
    n.breaks = 7
  ) +
  theme_bw()
```

Next we'll replicate Figure 3a from @antiaRoleEvolutionEmergence2003. This uses an extension of the 2-type (or one-step) branching process model to include multiple intermediate mutants/variants between the introduced wild-type and the fully-evolved pathogen with an $R > 1$. In @antiaRoleEvolutionEmergence2003 this is called the _jackpot model_. The introduced pathogen is subcritical ($R < 1$), and the intermediate variants have the same reproduction number as the wild-type. Only the final fully-evolved variant is supercritical ($R > 1$). 

We use the same number of intermediate mutants between the wild-type and the fully-evolved strain as @antiaRoleEvolutionEmergence2003.

```{r calc-prob-emerge-multi-type}
R_wild <- seq(0, 1.2, by = 0.01)
R_mutant <- 1.5
mutation_rate <- c(10^-1)
num_mutants <- 0:4

params <- expand.grid(
  R_wild = R_wild,
  R_mutant = R_mutant,
  mutation_rate = mutation_rate,
  num_mutants = num_mutants
)

prob_emerge <- apply(
  params,
  MARGIN = 1,
  FUN = function(x) {
    probability_emergence(
      R_wild = x[["R_wild"]],
      R_mutant = c(rep(x[["R_wild"]], x[["num_mutants"]]), x[["R_mutant"]]),
      mutation_rate = x[["mutation_rate"]],
      num_init_infect = 1
    )
  }
)
res <- cbind(params, prob_emerge = prob_emerge)
```

```{r plot-prob-emerge-multi-type}
#| fig.alt: >
#|   Line plot showing the probability of emergence of a pathogen as a function
#|   of the basic reproduction number (R0) of the introduced pathogen, with
#|   varying numbers of intermediate mutants before reaching the fully-evolved
#|   mutant. The x-axis represents the R0 of the introduced pathogen (ranging
#|   from 0 to 1.2), and the y-axis shows the probability of emergence on a
#|   logarithmic scale from 1e-05 to 1. Lines represent different numbers of
#|   intermediate mutants: red for 0, blue for 1, green for 2, purple for 3,
#|   and orange for 4. As the number of required intermediate mutants
#|   increases, the probability of emergence decreases across all R0 values.
#|   All curves rise with increasing R0, and converge at higher R0 values near
#|   1.2.
ggplot(data = res) +
  geom_line(
    mapping = aes(
      x = R_wild,
      y = prob_emerge,
      colour = as.factor(num_mutants)
    )
  ) +
  scale_y_continuous(
    name = "Probability of emergence",
    transform = "log",
    breaks = log_breaks(n = 6),
    limits = c(10^-5, 1)
  ) +
  scale_colour_brewer(
    name = "Number of \nintermediate mutants",
    palette = "Set1"
  ) +
  scale_x_continuous(
    name = expression(R[0] ~ of ~ introduced ~ pathogen),
    limits = c(0, 1.2),
    n.breaks = 7
  ) +
  theme_bw()
```

Thus far we've only introduced a single infection, however, just as with `probability_epidemic()` and `probability_extinct()`, we extend the method of @antiaRoleEvolutionEmergence2003 and introduce multiple initial human infections using the `num_init_infect` argument. 

```{r calc-prob-emerge-multi-init-infect}
R_wild <- seq(0, 1.2, by = 0.01)
R_mutant <- 1.5
mutation_rate <- 10^-1
num_mutants <- 1
num_init_infect <- seq(1, 9, by = 2)

params <- expand.grid(
  R_wild = R_wild,
  R_mutant = R_mutant,
  mutation_rate = mutation_rate,
  num_mutants = num_mutants,
  num_init_infect = num_init_infect
)

prob_emerge <- apply(
  params,
  MARGIN = 1,
  FUN = function(x) {
    probability_emergence(
      R_wild = x[["R_wild"]],
      R_mutant = c(rep(x[["R_wild"]], x[["num_mutants"]]), x[["R_mutant"]]),
      mutation_rate = x[["mutation_rate"]],
      num_init_infect = x[["num_init_infect"]]
    )
  }
)
res <- cbind(params, prob_emerge = prob_emerge)
```

```{r plot-prob-emerge-multi-init-infect}
#| fig.alt: >
#|   Line graph showing the relationship between the basic reproduction number
#|   (R0) of an introduced pathogen and the probability of emergence, under
#|   varying numbers of introductions. The x-axis represents R0 values from 0
#|   to 1.2, and the y-axis (logarithmic scale) represents the probability of
#|   emergence, from 1e-05 to 1. Five curves are shown for different numbers
#|   of introductions: 1 (red), 3 (blue), 5 (green), 7 (purple), and 9
#|   (orange). All curves show increasing probability of emergence with
#|   higher R0, and higher numbers of introductions consistently increase
#|   emergence probability. Increasing the number of initial infections has 
#|   diminishing increases on the probability of emergence with greater number
#|   of introductions.
ggplot(data = res) +
  geom_line(
    mapping = aes(
      x = R_wild,
      y = prob_emerge,
      colour = as.factor(num_init_infect)
    )
  ) +
  scale_y_continuous(
    name = "Probability of emergence",
    transform = "log",
    breaks = log_breaks(n = 6),
    limits = c(10^-5, 1)
  ) +
  scale_colour_brewer(
    name = "Number of introductions",
    palette = "Set1"
  ) +
  scale_x_continuous(
    name = expression(R[0] ~ of ~ introduced ~ pathogen),
    limits = c(0, 1.2),
    n.breaks = 7
  ) +
  theme_bw()
```
