---
title: "Pathogen evolution in the emergence of infectious disease outbreaks"
output: rmarkdown::html_vignette
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pathogen evolution in the emergence of infectious disease outbreaks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Infectious diseases that cannot transmit effectively between humans cannot cause sustained disease outbreaks. However, they can cause stuttering chains of transmission with several cases. These short human-to-human transmission chains provide opportunities for the pathogen to adapt to humans and enhance transmission potential. Therefore, even in cases where a pathogen initially introduced into a human population might have a reproduction number below 1, that is on average it is transmitted to less than one other person (bound to inevitably going extinct without pathogen evolution), if a pathogen mutates while infecting humans, its reproduction number might evolve to exceed 1, and emerge to cause sustained outbreaks. 

This vignette explores the `probability_emergence()` function. This is based on the framework of @antiaRoleEvolutionEmergence2003 which uses a multi-type branching process to calculate the probability that a subcritical zoonotic spillover will evolve into a more transmissible pathogen with a reproduction number greater than 1 and thus be able to cause a sustained epidemic from human-to-human transmission. 

```{r setup}
library(superspreading)
library(ggplot2)
library(scales)
```

We use `probability_emergence()` to replicate Figure 2b from @antiaRoleEvolutionEmergence2003 showing how the probability of emergence changes for different wild-type reproduction numbers and different mutation rates. This assumes a 2-type branching process with a wild-type pathogen with $R_0 < 1$ and a mutant pathogen with $R_0 > 1$ ($R_{\text{mutant}} \in {1.2, 1.5, 1,000}$).

```{r calc-prob-emerge}
R_wild <- seq(0, 1.2, by = 0.01)
R_mutant <- c( 1.2, 1.5, 1000)
mutation_rate <- c(10^-1, 10^-3)

params <- expand.grid(
  R_wild = R_wild, 
  R_mutant = R_mutant, 
  mutation_rate = mutation_rate
)

prob_emerge <- apply(
  params, 
  MARGIN = 1, 
  FUN = function(x) {
    probability_emergence(
      R_wild = x[["R_wild"]], 
      R_mutant = x[["R_mutant"]], 
      mutation_rate = x[["mutation_rate"]]
    )
  } 
)
res <- cbind(params, prob_emerge = prob_emerge)
```

```{r plot-prob-emerge, fig.width = 8, fig.height = 5}
ggplot(data = res) +
  geom_line(
    mapping = aes(
      x = R_wild, 
      y = prob_emerge, 
      colour = as.factor(mutation_rate), 
      linetype = as.factor(R_mutant)
    )
  ) +
  scale_y_continuous(
    name = "Probability of emergence",
    transform = "log",
    breaks = scales::log_breaks(n = 6),
    limits = c(10^-5, 1)
  ) +
  scale_colour_manual(
    name = "Mutation rate", 
    values = c("#228B22", "black")
  ) +
  scale_linetype_manual(
    name = "Mutant pathogen R0", 
    values = c("dotted", "dashed", "solid")
  ) +
  scale_x_continuous(
    name = expression(R[0]~of~introduced~pathogen), 
    limits = c(0, 1.2),
    n.breaks = 7
  ) +
  theme_bw()
```

Next we'll replicate Figure 3a from @antiaRoleEvolutionEmergence2003. This uses an extension of the 2-type (or one-step) branching process model to include multiple intermediate mutants/variants between the introduced wild-type and the fully-evolved pathogen with an $R > 1$. In @antiaRoleEvolutionEmergence2003 this is called the _jackpot model_. The introduced pathogen is subcritical ($R < 1$), and the intermediate variants have the same reproduction number as the wild-type. Only the final fully-evolved variant is supercritical ($R > 1$). 

We use the same number of intermediate mutants between the wild-type and the fully-evolved strain as @antiaRoleEvolutionEmergence2003.

```{r calc-prob-emerge-multi-type}
R_wild <- seq(0, 1.2, by = 0.01)
R_mutant <- c(1.5)
mutation_rate <- c(10^-1)
num_mutants <- 0:4

params <- expand.grid(
  R_wild = R_wild, 
  R_mutant = R_mutant, 
  mutation_rate = mutation_rate,
  num_mutants = num_mutants
)

prob_emerge <- apply(
  params, 
  MARGIN = 1, 
  FUN = function(x) {
    probability_emergence(
      R_wild = x[["R_wild"]], 
      R_mutant = c(rep(x[["R_wild"]], x[["num_mutants"]]), x[["R_mutant"]]), 
      mutation_rate = x[["mutation_rate"]]
    )
  } 
)
res <- cbind(params, prob_emerge = prob_emerge)
```

```{r plot-prob-emerge-multi-type, fig.width = 8, fig.height = 5}
ggplot(data = res) +
  geom_line(
    mapping = aes(
      x = R_wild, 
      y = prob_emerge, 
      colour = as.factor(num_mutants)
    )
  ) +
  scale_y_continuous(
    name = "Probability of emergence",
    transform = "log",
    breaks = scales::log_breaks(n = 6),
    limits = c(10^-5, 1)
  ) +
  scale_colour_brewer(
    name = "Number of \nintermediate mutants", 
    palette = "Set1"
  ) +
  scale_x_continuous(
    name = expression(R[0]~of~introduced~pathogen), 
    limits = c(0, 1.2),
    n.breaks = 7
  ) +
  theme_bw()
```
