---
title: "Epidemic Risk"
subtitle: "Superspreading on policy, decision making and control measures"
output: 
  bookdown::html_vignette2:
    code_folding: show
pkgdown:
   as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Epidemic Risk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

During the early stages of an infectious disease outbreak information on transmission
clusters can provide insight into the occurrence of superspreading events and from this
the probability an outbreak will cause an epidemic. This has implications for policy
decisions whether control interventions are required to bring disease transmission 
under control (e.g. $R$ < 1) and increase the probability that an outbreak goes extinct.

A recent example of where individual-level transmission and superspreading was analysed
to understand disease transmission was by the [Scientific Advisory Group for Emergencies 
(SAGE)](https://www.gov.uk/government/organisations/scientific-advisory-group-for-emergencies) 
for the [UK's COVID-19 response](https://www.gov.uk/government/publications/analysis-of-sars-cov-2-transmission-clusters-and-superspreading-events-3-june-2020). 

This vignette explores the applications of the functions included the {superspreading} package and their visualiation in understanding and informing decision makers for a variety of outbreak scenarios.

```{r setup}
library(superspreading)
library(ggplot2)
library(ggtext)
```

Early in the COVID-19 epidemic @kucharskiEarlyDynamicsTransmission2020 investigated
whether data on SARS-CoV-2 transmission displayed signatures of overdispersion (i.e.
evidence of superspreading events). This was in light of evidence that other corona virus
outbreaks (Severe acute respiratory syndrome (SARS) and Middle East respiratory syndrome (MERS))
had superspreading events. The detection of variability in individual-transmission
is important in understanding or predicting the likelihood of transmission establishing
when imported into a new location (e.g. country). 

```{r, prep-dispersion-plot}
epidemic_params <- expand.grid(
  R = 2.35, R_lw = 1.5, R_up = 3.2, k = seq(0, 1, 0.1), a = seq(0, 10, 1)
)
# results are transposed to pivot to long rather than wide data
prob_epidemic <- t(apply(epidemic_params, 1, function(x) {
  median <- probability_epidemic(R = x[["R"]], k = x[["k"]], a = x[["a"]])
  lower <- probability_epidemic(R = x[["R_lw"]], k = x[["k"]], a = x[["a"]])
  upper <- probability_epidemic(R = x[["R_up"]], k = x[["k"]], a = x[["a"]])
  c(prob_epidemic = median, prob_epidemic_lw = lower, prob_epidemic_up = upper)
}))
epidemic_params <- cbind(epidemic_params, prob_epidemic)

# subset data for a single initial infection
homogeneity <- subset(epidemic_params, a == 1)
```

```{r, plot-dispersion, class.source = 'fold-hide', fig.cap="The probability that an initial infection (introduction) will cause a sustained outbreak (transmission chain). The dispersion of the individual-level transmission is plotted on the x-axis and probability of outbreak -- calculated using `probability_epidemic()` -- is on the y-axis. This plot is reproduced from Kucharski et al. (2020) figure 3A.", fig.width = 8, fig.height = 5}
# plot probability of epidemic across dispersion
ggplot(data = homogeneity) +
  geom_ribbon(
    mapping = aes(
      x = k,
      ymin = prob_epidemic_lw,
      ymax = prob_epidemic_up
    ),
    fill = "grey70"
  ) +
  geom_line(mapping = aes(x = k, y = prob_epidemic)) +
  geom_vline(
    mapping = aes(xintercept = 0.2),
    linetype = "dashed"
  ) +
  annotate(geom = "text", x = 0.15, y = 0.75, label = "SARS") +
  geom_vline(
    mapping = aes(xintercept = 0.4),
    linetype = "dashed"
  ) +
  annotate(geom = "text", x = 0.45, y = 0.75, label = "MERS") +
  scale_y_continuous(
    name = "Probability of large outbreak",
    limits = c(0, 1)
  ) +
  scale_x_continuous(name = "Extent of homogeneity in transmission") +
  theme_bw()
```

The degree of variability in transmission increases as $k$ decreases. So the probability
of a large outbreak is smaller for smaller values of $k$, meaning that if COVID-19
is more similar to SARS than MERS it will be less likely to establish and cause an 
outbreak if introduced into a newly susceptible population.

However, this probability of establishment and continued transmission will also
depend on the number of initial introductions to that populations. The more
introductions, the higher the chance one will lead to an epidemic.

```{r, prep-introductions-plot}
introductions <- subset(epidemic_params, k == 0.5)
```

```{r, plot-introductions, class.source = 'fold-hide', fig.cap="The probability that an a number of introduction events will cause a sustained outbreak (transmission chain). The number of disease introductions is plotted on the x-axis and probability of outbreak -- calculated using `probability_epidemic()` -- is on the y-axis. This plot is reproduced from Kucharski et al. (2020) figure 3B.", fig.width = 8, fig.height = 5}
# plot probability of epidemic across introductions
ggplot(data = introductions) +
  geom_pointrange(
    mapping = aes(
      x = a,
      y = prob_epidemic,
      ymin = prob_epidemic_lw,
      ymax = prob_epidemic_up
    )
  ) +
  scale_y_continuous(
    name = "Probability of large outbreak",
    limits = c(0, 1)
  ) +
  scale_x_continuous(name = "Number of introductions", n.breaks = 6) +
  theme_bw()
```

Different levels of heterogeneity in transmission will produce different probabilities of epidemics.

```{r, plot-introductions-multi-k, class.source = 'fold-hide', fig.cap="The probability that an a number of introduction events will cause a sustained outbreak (transmission chain). The number of disease introductions is plotted on the x-axis and probability of outbreak -- calculated using `probability_epidemic()` -- is on the y-axis. Different values of dispersion are plotted to show the effect of increased transmission variability on an epidemic establishing", fig.width = 8, fig.height = 5}
# plot probability of epidemic across introductions for multiple k
ggplot(data = epidemic_params) +
  geom_point(
    mapping = aes(
      x = a,
      y = prob_epidemic,
      colour = k
    )
  ) +
  scale_y_continuous(
    name = "Probability of large outbreak",
    limits = c(0, 1)
  ) +
  labs(colour = "Dispersion (*k*)") +
  scale_x_continuous(name = "Number of introductions", n.breaks = 6) +
  scale_colour_continuous(type = "viridis") +
  theme_bw() +
  theme(legend.title = element_markdown())
```

In the above plot we drop the uncertainty around each point and assume a known value
of $R$ in order to more clearly show the pattern.

Applications that make it easy to explore this functionality and compare between existing
parameter estimates of offspring distributions are also useful. An example of this is the
[Shiny app developed by the Centre for Mathematical Modelling of Infectious diseases](https://cmmid.github.io/visualisations/new-outbreak-probability) at the London
School of Hygiene and Tropical Medicine, for comparing SARS-like and MERS-like scenarios, 
as well as random mixing.

Conversely to the probability of an epidemics, the probability that an outbreak will 
go extinct (i.e. transmission will subside), can also be plotted for different values of 
$R$.

```{r, prep-exinction-plot}
extinction_params <- expand.grid(
  R = seq(0, 5, 0.1),
  k = c(0.01, 0.1, 0.5, 1, 4, Inf),
  a = 1
)
# results are transposed to pivot to long rather than wide data
prob_extinct <- apply(extinction_params, 1, function(x) {
  median <- probability_extinct(R = x[["R"]], k = x[["k"]], a = x[["a"]])
  median
})
extinction_params <- cbind(extinction_params, prob_extinct)
```

```{r, plot-extinction, class.source = 'fold-hide', fig.cap="The probability that an infectious disease will go extinct for a given value of $R$ and $k$. This is calculated using `probability_extinct()` function. This plot is reproduced from @lloyd-smithSuperspreadingEffectIndividual2005 figure 2B.", fig.width = 8, fig.height = 5}
# plot probability of extinction across R for multiple k
ggplot(data = extinction_params) +
  geom_point(
    mapping = aes(
      x = R,
      y = prob_extinct,
      colour = factor(k)
    )
  ) +
  scale_y_continuous(
    name = "Probability of extinction",
    limits = c(0, 1)
  ) +
  labs(colour = "Dispersion (*k*)") +
  scale_x_continuous(
    name = "Reproductive number (*R*)",
    n.breaks = 6
  ) +
  scale_color_viridis_d() +
  theme_bw() +
  theme(
    axis.title.x = element_markdown(),
    legend.title = element_markdown()
  )
```

## Superspreading in decision making

One of the first tasks in an outbreak is to establish whether estimates of individual-level
transmission variability have been reported, either for this outbreak or a previous outbreak
of the same pathogen. Alternatively, if it is an outbreak of a novel pathogen, parameters
for similar pathogens can be searched for. 

This is accomplished by another Epiverse-TRACE R package: [{epiparameter}](https://epiverse-trace.github.io/epiparameter/index.html). This package
contains a library of epidemiological parameters included offspring distribution parameter estimates and uncertainty obtained from published literature. This aids the collation and establishes a baseline understanding of transmission. Similar to the collation of transmission clusters for [COVID-19](https://www.gov.uk/government/publications/analysis-of-sars-cov-2-transmission-clusters-and-superspreading-events-3-june-2020).

Equipped with these parameter estimates, metrics that provide an intuitive understanding of transmission patterns can be tabulated.
For example the proportion of transmission events in a given cluster size can be calculated using the {superspreading} function `proportion_cluster_size()`.

```{r}
# For R = 0.8
proportion_cluster_size(
  R = 0.8,
  k = seq(0.1, 1, 0.1),
  cluster_size = c(5, 10, 25)
)

# For R = 3
proportion_cluster_size(
  R = 0.8,
  k = seq(0.1, 1, 0.1),
  cluster_size = c(5, 10, 25)
)
```

These results indicate whether preventing gatherings of a certain size can reduce the epidemic
by preventing potential superspreading events.

When data on the settings of transmission is known, priority can be given to restricting 
particular types of gatherings to most effectively bring down the reproduction number.

## Controls on transmission

In their work on identifying the prevalence of overdispersion in individual-level transmission, @lloyd-smithSuperspreadingEffectIndividual2005 also showed the effect of control measures on the probability of containment. They defined containment as the size of a transmission chain not reaching 100 cases. They assumed control measures could be either population-level, or individual-level. {superspreading} provides `probability_contain()` to calculate the probability that disease spread will go extinct before reaching a threshold size.

```{r, prep-containment-plot}
contain_params <- expand.grid(
  R = 3, k = c(0.1, 0.5, 1, Inf), a = 1, c = seq(0, 1, 0.05)
)
prob_contain <- apply(contain_params, 1, function(x) {
  probability_contain(
    R = x[["R"]],
    k = x[["k"]],
    a = x[["a"]],
    c = x[["c"]],
    control_type = "population",
    stochastic = TRUE,
    case_threshold = 100
  )
})
contain_params <- cbind(contain_params, prob_contain)
```

```{r, plot-containment, class.source = 'fold-hide', fig.cap="The probability that an outbreak will be contained (i.e. not exceed 100 cases) for a variety of population-level control measures. The probability of containment is calculated using `probability_contain()`. This plot is reproduced from Lloyd-Smith et al. (2005) figure 3C.", fig.width = 8, fig.height = 5}
# plot probability of epidemic across introductions for multiple k
ggplot(data = contain_params) +
  geom_point(
    mapping = aes(
      x = c,
      y = prob_contain,
      colour = factor(k)
    )
  ) +
  scale_y_continuous(
    name = "Probability of containment",
    limits = c(0, 1)
  ) +
  scale_x_continuous(name = "Control measures (*c*)", n.breaks = 6) +
  labs(colour = "Dispersion (*k*)") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(
    axis.title.x = element_markdown(),
    legend.title = element_markdown()
  )
```
