---
title: "Getting started with {superspreading}"
output: 
  bookdown::html_vignette2:
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting started with {superspreading}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The {superspreading} R package provides a set of functions for understanding
individual-level tranmsission dynamics, and thus whether there is evidence of
superspreading or superspreading events (SSE).

Superspreading is the heterogeneity in disease transmission, beyond 
stochasticity, such that some individuals transmit to many infectees while other
infectors infect few or zero individuals @lloyd-smithSuperspreadingEffectIndividual2005.

```{r setup}
library(superspreading)
library(epiparameter)
```

## Probability of epidemic

The probability that a novel disease will cause a epidemic (i.e. sustained 
transmission in the population) is determined by the nature of that diseases
transmission heterogeneity. This variability may be an intrinsic property of 
the disease, or a product of human behaviour and social mixing patterns.

For a given value of R, that is the number of secondary cases caused by a 
primary case, if the variability is high, the probability that the outbreak 
will cause epidemic is low as the superspreading events are rare. Whereas for
lower variability the probability is higher as more individuals are closer to the 
mean (R). The `probability_epidemic()` function in {superspreading} can calculate
this probability. `k` is the variability.

```{r, prob-epidemic-k}
probability_epidemic(R = 1.5, k = 1, a = 1)
probability_epidemic(R = 1.5, k = 0.5, a = 1)
probability_epidemic(R = 1.5, k = 0.1, a = 1)
```

In the above code `k` is the overdispersion parameter of a negative binomial
distribution. Values above to one are low heterogeneity (in the case `k` 
&rarr; $\infty$ it is a poisson distribution), and as it decreases to zero 
heterogeneity increases. When `k` equals 1 the distribution geometric.

When the value of R increases, this inevitably causes the probability of an 
epidemic to increase.

```{r, prob-epidemic-R}
probability_epidemic(R = 0.5, k = 1, a = 1)
probability_epidemic(R = 1.0, k = 1, a = 1)
probability_epidemic(R = 1.5, k = 1, a = 1)
probability_epidemic(R = 5, k = 1, a = 1)
```

Any value of R less than or equal to one will have zero probability of causing
a sustained epidemic.

Lastly on the probability of transmission dynamics causing an epidemic, the 
number of initial infections seeding the outbreak will influence the probability.
In `probability_epidemic()` `a` is the number of initial infections.

```{r, prob-epidemic-a}
probability_epidemic(R = 1.5, k = 1, a = 1)
probability_epidemic(R = 1.5, k = 1, a = 10)
probability_epidemic(R = 1.5, k = 1, a = 100)
```

## References
